old snippets

### 2024-12-18 manuscript/manuscript.qmd
updated manuscript .qmd

---
bibliography: references.bib
csl: apa.csl
format: 
  docx:
    reference-doc: reference-doc.docx
output:
  officedown::rdocx_document:
    page_margins:
      bottom: 1
      footer: 0
      gutter: 0
      header: 0.5
      left: 1
      right: 1
      top: 1
    plots:
      align: center
      caption:
        pre: 'Figure '
        sep: '. '
        style: Image Caption
    tables:
      caption:
        pre: 'Table '
        sep: '. '
        style: Table Caption
  pdf_document: default
  word_document: default
editor: 
  markdown: 
    wrap: sentence
---

@@

@@

#### **Attention shifting and inhibition abilities among people living in adverse conditions**

@@

# Corresponding author

@@

# Data Availability

All scripts and materials needed to reproduce the findings are available on the article's Github repository ([https://stefanvermeent.github.io/liss_wm_profiles_2023/](https://stefanvermeent.github.io/liss_wm_profiles_2023/)).
We also include instructions on how to reproduce each step of our analyses. 
In this paper, we make use of data from the LISS panel (Longitudinal Internet studies for the Social Sciences) managed by the non-profit research institute Centerdata (Tilburg University, the Netherlands). 
All datasets are available in the LISS data archive. 
Researchers who want to access the data are required to sign a statement confirming that information about individual persons, households, etc., will not be released to others (go to [https://statements.centerdata.nl](https://statements.centerdata.nl) for more information).

# Funding statement
This research was conducted in whole or in part using ODISSEI, the Open Data Infrastructure for Social Science and Economic Innovations(<https://ror.org/03m8v6t10>).
WEF’s contributions have been supported by the Dutch Research Council (V1.Vidi.195.130) and the James S. McDonnell Foundation (https://doi.org/10.37717/220020502).

# Disclosures

We declare no conflicts of interest.

# Ethics Approval Statement

This study was approved by the Ethics Review Board of the Faculty of Social & Behavioural Sciences of Utrecht University (FETC20-490) and the Ethics committee for research in the Sciences and Life Sciences of the University of Amsterdam (FNWI-41_2023).

\pagebreak

Exposure to adversity (i.e., prolonged exposure to intense stress) is associated with robust, persisting changes in executive functioning (EF) [@ellis_2022; @farah_2006; @sheridan_2022; @sheridan_2014].
Given the importance of EF---a set of abilities involved in the top-down, goal-directed regulation of behavior [@miyake_2000; @rey_mermet_2019]---much research has been devoted to understanding which EF abilities are most strongly associated with adversity.
In some cases, these insights have already been used to screen for early EF deficits, and to develop interventions to help bridge performance gaps [@distefano_2021].

A lot of what we know about the association between adversity and EF stems from analyzing raw performance, most often response times (RT) or accuracy rates.
In addition, it is common for studies to only include a single task per EF ability.
For example, previous studies have inferred difficulties in ignoring distractions (inhibition) in people from adversity using a single inhibition task [e.g., @farah_2006; @fields_2021; @mezzacappa_2004; @noble_2005].
The same thing is true for observed adaptive benefits in the ability to shift attention between tasks (attention shifting) [@fields_2021; @howard_2020; @nweze_2021; @young_2022; @mittal_2015].
In effect, an important assumption underlying many deficit and adaptation findings is that raw task performance offers a reliable, relatively pure estimate of the EF ability [@white_2022].

Unfortunately, an increasing number of psychometric investigations into EF do not support this assumption [@krumm_2009; @loffler_2024; @rey_mermet_2019; @rouder_2019; @stahl_2014].
A central issue is that raw task performance is not task-pure.
Aside from the EF ability needed to solve the task, a person's RT is also influenced by other factors such as response caution (i.e., their relative emphasis on speed versus accuracy).
Accounting for these factors using cognitive modeling, a recent study found that children with more household threat were slower on EF tasks partly because they were more cautious [@vermeent_2024a].

Aside from different processing stages, performance across EF tasks is also influenced by general processing speed [@frischkorn_2019; @lerche_2020; @loffler_2024].

[Might not be any ]
Even more problematic are recent studies suggesting that performance on EF tasks taps into little more than general processing speed.
For example, after accounting for other factors such as response caution, another study found that a common (latent) EF factor was fully explained by performance on really basic processing speed tasks [@loffler_2024].
A few studies have focused on this distinction in the context of adversity and/or socioeconomical status.
on a raw performance level, lower SES was found to be negatively correlated with a task-general factor, with additional associations on the task-specific level.
In another

# # Implications and potential solutions for adversity research

An important implication of this work is that standard assessments---based on raw performance of single EF tasks---may be ill-suited at estimating associations between adversity and specific EF abilities.
Patterns that may look like deficits (or adaptations) in specific abilities may not be driven by these abilities at all.
For example, if adversity is associated with slower task-general processing speed, analyses focused on individual EF tasks could make it seem like many different abilities impaired, instead of a single, general process.

Fortunately, these challenges can be tackled using a combination of cognitive modeling and structural equation modeling.
As a first step, ability-relevant processing should be separated from ability-irrelevant processes, such as response caution or the time taken to prepare for a task [...].
One effective way of doing this is through the use of cognitive models like the Drift Diffusion Model [DDM\; @forstmann_2016; @ratcliff_1998; @ratcliff_2008; @wagenmakers_2009].
The DDM broadly divides the time taken on speeded EF tasks into a decision-making stage and a non-decision-making stage [see Figure 1].
In the decision-making stage, people accumulate evidence for one of two responses (e.g., left or right button), until they have enough evidence for one response over the other.
The amount of evidence a person requires can vary between people (as well as between tasks within the same person) [XXX], and is modeled as the distance between the decision boundaries for each respective response.
Finally, the DDM models two non-decision stages: one at the start of the trial (e.g., task preparations, stimulus encoding) and one at the end of the trial (i.e., response execution).

When applied to empirical performance data, the DDM distinguishes between three parameters mapping onto distinct cognitive processes.
The *drift rate* reflects the rate at which the evidence accumulation process drifts towards the decision boundary, and thus provides a measure of the speed/efficiency of information processing.
The *boundary separation* reflects the width between the two decision boundaries, and provides a measure of a person's response caution.
Finally, the *non-decision time* reflects a combination of preparation time (e.g., stimulus encoding) as well as execution time (e.g., time spent pressing the button).

Researchers have argued that the drift rate on EF tasks is a more valid measure of EF ability, at least to the extent to which the tasks are reliable measures of EF ability [@loffler_2024; @vermeent_2024a; @weigard_2021].
However, drift rates are still influenced at least partly by general speed of processing.
Theoretically, this should be the part of the variance that is shared across drift rates of all tasks, and of conditions both including an EF component (e.g., incongruent trials) and excluding an EF component (e.g., congruent trials) [@loffler_2024].
This part of the variance can be estimated on a latent level using SEM.
Any remaining task-specific variance then contains EF-specific variance.

As mentioned above, a recent investigation did not find significant associations between adversity and task-specific EF variances [@vermeent_2024a].
This could mean one of two things.
First, task-specific variance may reliably reflect specific EF abilities, and adversity may not be associated with these abilities.
Second, task-specific variance may *not* reliably reflect specific EF abilities.
That is, after accounting for general speed of processing, little more than measurement error may remain.
Testing these hypotheses requires sampling two or more tasks that are supposed to measure the same ability, and using latent modeling to investigate whether they form a consistent latent factor after accounting for general speed of processing.

@@

@@

Participants were 1,000 Dutch people randomly drawn from the Longitudinal Internet studies for the Social Sciences (LISS) panel [@scherpenzeel_2011].
The LISS panel is an invite-only, representative probability sample of the Dutch population consisting of roughly 7,500 individuals across 5,000 households.
LISS participants complete a yearly core battery of questionnaires about various key domains, among which their financial situation over the past year.
In addition, they can participate in further monthly studies covering a wide range of different topics.
The current study was fielded between May and June 2024.
Participants were able to complete the cognitive tasks across two or more sessions to increase participation rates.
People were eligible for participation if they (1) were between 18 and 55 years old; (2) if they had completed at least one wave of a study in the LISS archive assessing exposure to crime (see below); (3) if they had agreed to linking their LISS data to government microdata (not relevant here).
Participants were excluded if they interacted with other browser tabs mid-trial on any of the cognitive tasks (but not during breaks or in-between tasks). 
If performance on a specific task was at or below chance level, we excluded the data for that task but retained other task data of that participant.

@@

We preregistered our approach to compute adversity composites based on observed correlations between measures.
In short, if all measures of an adversity type correlated .60 or higher (indicating "strong" correlations), we calculated a uniformly weighted average.
If one or more correlations were lower than .60, we used Principal Component Analysis (PCA) to the separate measures and extracted only the first principal component score.
Perceived scarcity and income-to-needs were considered as two separate types of material deprivation (and unpredictability), regardless of their correlation.

### Neighborhood threat

*Perceived neighborhood crime.* Participants completed the Neighborhood Violence Scale [NVS\; @frankenhuis_2018; @frankenhuis_2020] in the current study.
The NVS included seven items about current perceived neighborhood threat (e.g., “Where I live, it is important to be able to defend yourself against physical harm”), on a scale of 1 ("Completely disagree") to 7 ("Completely agree"). 
All items of the NVS were averaged together and standardized.
We also included four items on perceived neighborhood threat from the LISS archive (six waves: <https://doi.org/10.17026/dans-zch-j8xt>), in which participant reported how often it happens that they 1) “avoid certain areas in your place of residence because you perceive them as unsafe”, 2) “do not respond to a call at the door because you feel that it is unsafe”, 3) “leave valuable items at home to avoid theft or robbery in the street?”, 4) “make a detour, by car or on foot, to avoid unsafe areas?” on a scale of 1 (“(Almost) never”), 2 (“Sometimes”), or 3 (“Often”). 
We recoded these items so that "(Almost) never" corresponded to zero.
We summed the items within each wave, and then calculated an average across waves for which participants had data.

*Crime victimization.* We used seven items from the LISS archive (six waves: <https://doi.org/10.17026/dans-zch-j8xt>) in which participants reported whether or not they were the victim of eight types of crime in the last two years: (1) burglary or attempted burglary; (2) theft from their car; (3) theft of their wallet or purse, handbag, or other personal possession; (4) wreckage of their car or other private property; (5) intimidation by any other means; (6) maltreatment of such serious nature that it required medical attention; (7) maltreatment that did not require medical attention. 
As the last wave occurred in 2018, we included the same items in the current study asking about people's crime exposure in the last two years.
We computed the total number of distinct crimes that participants were exposed to at any moment in time (a 'variety score'\; @sweeten_2012).

*Neighborhood threat composite.* Correlations among the two measures of perceived neighborhood crime and crime victimization were low.
Therefore, we used PCA to extract the first principal component score.
This score accounted for XXX% of the variance in the three measures.

### Material deprivation

*Perceived scarcity (mean).* Measures of perceived scarcity were derived from the LISS archive, using the yearly recurring core study on household and personal income (16 waves: <https://doi.org/10.57990/1gr4-bf42>).
First, participants reported how difficult it currently is to live off the income of their household, on a scale from 0 (very hard) to 10 (very easy). 
Second, participants reported which of the following best applied to their current financial situation: (1) “we are accumulating debt”; (2) “we are somewhat eating into savings”; (3) “we are just managing to make ends meet”; (4) “we have a little bit of money to spare”; (5) “we have a lot of money to spare”. 
Third, participants reported which of the following applied to their current financial situation (0 = no, 1 = yes): (1) “having trouble making ends meet”; (2) unable to quickly replace things that break”; (3) “having to lend money for necessary expenditures”; (4) “running behind in paying rent/mortgage or general utilities”; (5) “debt collector/bailiff at the door in the last month”; (6) “received financial support from family or friends in the last month”.
Responses were coded so that higher scores indicated more perceived scarcity.

We first averaged each measure separately across waves, and then scaled them.
As all item correlations were > .60, we computed a uniformly weighted average.

*Income-to-needs (mean).* The income-to-needs ratio was derived from monthly self-reported net household income in the LISS archive (<https://doi.org/10.57990/qn3k-as78>).
Reported income of zero was set to missing, as this could indicate an unwillingness to report on income rather than a true income of zero.
If net household income data was missing (or zero) for an entire year, we imputed, if possible, the reported yearly net household income from the core study on income (<https://doi.org/10.57990/1gr4-bf42>), dividing by 12 to obtain monthly estimates.
We then divided the income by the poverty threshold set by Statistics Netherlands (CBS) [@brakel_2023; CBS, personal communication, December 15, 2023].
For households with more than three children, we used the threshold for a family with three children, as CBS does not calculate thresholds for families larger than three children.
For the same reason, income of households with more than two adults was compared to the threshold for households with two adults.
We computed the average income-to-needs ratio across years, reverse coding it so that higher scores indicated households further below the threshold.

### Unpredictability

*Perceived scarcity (SD/mean).* Unpredictability over time was computed using the coefficient of variation, which is the within-person standard deviation across years divided by the mean across years (@key_2017; @liu_2022; @ugarte_2023; @walasek_2024; @young_2020).
This measure accounts for the fact that the mean and standard deviation tend to be highly positively correlated.
Unpredictability in perceived scarcity was derived from the same measures as the mean level of perceived scarcity (see above).
We first computed the standard deviation for each measure separately across years, and then scaled them.
We then computed a uniformly weighted average, which was divided by the mean level of perceived scarcity across years.

*Income-to-needs (SD/mean).* Unpredictability over time in the income-to-needs ratio was also computed using the coefficient of variation.
We first computed the within-person standard deviation of the income-to-needs ratio across years, and then divided it by the mean income-to-needs ratio across years.

@@

All cognitive tasks were programmed in jsPsych 7.3 [@deleeuw_2015].
At the start of the session, participants entered fullscreen mode to avoid distractions from other browser tabs.
The tasks were presented against a light-gray background.

*Flanker Task.* This task measures inhibition of distractor interference [...]
On each trial, participants saw five arrows side-by-side horizontally, pointing either left or right.
Their task was to indicate the direction of the central arrow.
The arrows were randomly presented 300 pixels above or below the center of the screen.
On 50% of the trials, all arrows pointed in the same direction (congruent trials).
On the other half, the arrows surrounding the central arrow pointed in the opposite direction (incongruent trials).
Participants first completed eight practice trials, followed by two test blocks of 32 trials each, for a total of 64 trials.

>>>>

---
bibliography: references.bib
csl: apa.csl
format: 
  docx:
    reference-doc: reference-doc.docx
output:
  officedown::rdocx_document:
    page_margins:
      bottom: 1
      footer: 0
      gutter: 0
      header: 0.5
      left: 1
      right: 1
      top: 1
    plots:
      align: center
      caption:
        pre: 'Figure '
        sep: '. '
        style: Image Caption
    tables:
      caption:
        pre: 'Table '
        sep: '. '
        style: Table Caption
  pdf_document: default
  word_document: default
---

@@

source('../scripts/0_functions/general-functions.R')

load("../data/exclusions.RData")
load("../analysis_objects/n_sessions.RData")
load("../analysis_objects/results_confirmatory.RData")
load("../analysis_objects/results_exploratory.RData")
load("../analysis_objects/rendered_figures.RData")
load("../analysis_objects/rendered_tables.RData")

@@

#### **Adversity is associated with lower general processing speed rather than specific executive functioning abilities**

@@

# Author Note

This research was conducted in part using ODISSEI, the Open Data Infrastructure for Social Science and Economic Innovations (<https://ror.org/03m8v6t10>).
WEF’s contributions have been supported by the Dutch Research Council (V1.Vidi.195.130) and the James S. McDonnell Foundation (https://doi.org/10.37717/220020502).

Stefan Vermeent served as lead for conceptualization, data curation, methodology, formal analysis, and writing the original draft. Anna-Lena Schubert served in a supporting role for conceptualization, formal analysis, methodology, and writing-review and editing. Willem E. Frankenhuis served in a supporting role for conceptualization, methodology, supervision, funding acquisition, and writing-review and editing.

The preregistration, analysis scripts, and study materials are available on the article's Github repository ([https://stefanvermeent.github.io/liss_ef_2024/](https://stefanvermeent.github.io/liss_ef_2024/)).
In this paper, we make use of data from the LISS panel (Longitudinal Internet studies for the Social Sciences) managed by the non-profit research institute Centerdata (Tilburg University, the Netherlands). 
All datasets are available to researchers in the LISS data archive, after signing a data use agreement ([https://www.dataarchive.lissdata.nl/study-units/view/1](https://www.dataarchive.lissdata.nl/study-units/view/1)).

@@

We have no conflict of interest to disclose.

**word count:** 9573

{{< pagebreak >}}

# Abstract

Exposure to adversity may impair executive functioning (i.e., deficit frameworks), but could also enhance, or leave intact, specific EF abilities (i.e., adaptation frameworks).
Both frameworks often use raw performance (e.g., speed) to estimate EF ability. 
However, this approach (1) conflates different cognitive processes, and (2) generally does not distinguish specific EF abilities from processes that are shared across EF tasks, such as general processing speed.
Here, we integrate deficit and adaptation frameworks by building bridges with mathematical and cognitive psychology.
Specifically, we use cognitive modeling (Drift Diffusion Modeling) to isolate different cognitive processes: speed of information accumulation, response caution, and speed of stimulus encoding and response execution.
We then use structural equation modeling to investigate whether associations between adversity and cognitive processes are task-general or ability-specific.
We recruited 1061 participants from the Dutch LISS panel.
Participants completed a basic processing speed task, two inhibition tasks, and three attention-shifting tasks.
We measured exposure to threat and material deprivation in childhood and adulthood.
Exposure to threat (but not material deprivation) in adulthood was negatively associated with task-general processing speed.
After accounting for task-general processes, remaining variance was not related to either inhibition or attention-shifting ability.
Non-preregistered analyses showed that childhood exposure to material deprivation and threat were negatively associated with (1) general processing speed, and (2) task-specific information accumulation.
The latter reflected unique features of individual tasks, rather than specific EF abilities.
Taken together, these results suggest that adversity researchers overestimate associations between adversity and specific EF abilities when analyzing raw performance.

# Public significance statement

Decades of research suggests that adversity impairs executive functioning abilities, such as inhibition and attention shifting.
These conclusions are often derived from analyzing raw performance, such as response times or accuracy.
We use cognitive modeling to reveal that adversity lowers general processing speed, but not specific executive functioning abilities.
This suggests that adversity research likely overestimates the extent to which adversity is associated with executive functioning abilities.

*Keywords*: adversity, executive functioning, Drift Diffusion Modeling, structural equation modeling

{{< pagebreak >}}

Psychologists have used two main frameworks to explain associations between adversity exposure---such as material deprivation or threat---and executive functioning (EF); one focuses on deficits, the other on adaptation.
Deficit frameworks predict that adverse experiences impair brain structure and function in ways that undermine social and cognitive abilities [@farah_2006; @mertz_2019; @rosen_2018; @ursache_2016a].
Adaptation frameworks predict that adverse experiences can lead to the development of intact or enhanced EF abilities, specifically those abilities useful for solving challenges posed by adverse environments [@ellis_2022; @frankenhuis_2013; @frankenhuis_2020].

Although the two frameworks make opposing predictions in some cases, they share the goal to reveal how adversity influences specific EF abilities [@farah_2006; @ellis_2022; @frankenhuis_2020; @johnson_2021].
Knowledge about impaired abilities provides valuable targets for interventions designed to bridge achievement gaps.
Conversely, knowledge about enhanced abilities can be leveraged in school or work contexts, for instance, by redesigning learning settings to match people’s unique strengths.

Research within both frameworks typically employs the same inferential strategy, by estimating specific EF abilities based on raw performance on EF tasks (e.g., response speed or accuracy).
For instance, on the one hand, children who live in less favorable environments tend to be slower and less accurate on inhibition tasks---which has been interpreted as an impaired ability to inhibit distractions [@farah_2006; @fields_2021; @mezzacappa_2004; @mittal_2015; @noble_2005].
On the other hand, some studies report that adolescents and young adults with more exposure to threat and unpredictability might perform better at shifting their attention between different tasks, without sacrificing much performance on each task---which may be a useful adaptation for tracking information in fast changing and dangerous environments [@fields_2021; @howard_2020; @mittal_2015; @nweze_2021; @young_2022].
In this paper, we will challenge the inferential strategy of estimating EF abilities based on raw performance. 
We advocate for cognitive modeling as a viable alternative approach to estimating EF abilities.

# Methodological challenges for measuring specific EF abilities

Recent research shows that raw performance on EF tasks does not necessarily capture specific EF abilities, for two related reasons.
First, raw performance on EF tasks is also influenced by processes other than the ability of interest, such as response caution and the speed of response execution [@hedge_2021; @ratcliff_2008; @stahl_2014].
Thus, slower (or faster) responses in people with more adversity exposure need not reflect EF ability, but could, for instance, reflect a tendency to respond more cautiously [@vermeent_2024a].
Second, raw performance on EF tasks is also influenced by general processes, such as basic speed of information processing [@frischkorn_2019; @loffler_2024; @weigard_2021].
Thus, lower performance on multiple EF tasks may reflect a difference in a general process, rather than specific EF abilities [@bignardi_2024; @vermeent_2024a; @young_2024].

The standard inferential strategy of estimating EF abilities from raw performance limits the effectiveness of interventions. 
For example, some interventions include training modules for specific EF tasks if performance on a task is below some prespecified cut-off [@distefano_2021].
Performance on inhibition tasks in particular is thought to mediate the association between adversity exposure and achievement [@taylor_2022].
At face value, this would suggest interventions targeting inhibition ability (e.g., inhibition training, which might involve removing distractions from the environment) could positively impact achievement outcomes.
However, if impaired inhibition ability is *not* the root cause of performance differences, but rather general processes are (such as the speed of information processing), the impact of such interventions will likely be limited.
Therefore, adversity research should move beyond the use of raw performance to better understand the associations between adversity exposures and specific EF abilities.

# Separating EF ability from other decision-making processes

Raw performance confounds multiple stages of processing: from initial preparations (e.g., stimulus encoding), to processing task-relevant information, to deciding on and executing a response [@ratcliff_1978; @forstmann_2016; @wagenmakers_2009].
Individual differences in response times and error rates also depend on how cautiously people make decisions.
More cautious people tend to respond more slowly to increase their accuracy, which does not necessarily imply lower ability [@voss_2004].
Thus, understanding associations between adversity and specific EF abilities requires isolating ability-relevant processing (e.g., inhibition, attention-shifting) from ability-irrelevant processes that contribute to response times.

Cognitive modeling can distinguish between abilities and ability-irrelevant processes. 
Here, we focus on the Drift Diffusion Model [DDM\; @forstmann_2016; @ratcliff_1998; @ratcliff_2008; @wagenmakers_2009], which can be applied to many widely used EF tasks.
The DDM assumes that people accumulate evidence for one of two responses (e.g., left or right button) until they have acquired sufficient evidence to make a decision (see Figure 1).
When applied to trial-level response times and accuracy data, the DDM estimates four parameters that represent distinct cognitive processes [@voss_2004].
The *drift rate* reflects the rate at which people accumulate evidence for the correct response; thus, it measures the efficiency of information processing.
A higher drift rate leads to faster responses and higher accuracy.
When applied to EF tasks, the drift rate captures individual differences in specific EF abilities [although drift rate may also capture more general processes, see below\; @loffler_2024; @vermeent_2024a; @weigard_2021].
The *boundary separation* reflects the width between the two decision boundaries; it measures a person's response caution.
A higher boundary separation leads to slower responses and higher accuracy.
The *non-decision time* reflects a combination of preparation time (e.g., stimulus encoding) as well as execution time (e.g., time spent pressing the button).
A larger non-decision time leads to slower responses without a change in accuracy.
Finally, the *starting-point* allows for initial bias for one of two responses (e.g., happy or angry faces). 
In this paper, we will not consider the starting point, because people cannot have biases towards correct or incorrect responses.

```{r}
#| label: Figure1
#| fig-width: 6.5
#| fig-height: 7
#| dpi: 600
#| out-width: 6in
#| fig-cap: | 
#|   **Figure 1.** Visual representation of the Drift Diffusion Model (DDM). The DDM assumes that people accumulate evidence for one of two responses until they have acquired sufficient evidence to make a decision. Each squiggly line represents the evidence accumulation process on a single trial. The thick, horizontal lines represent the decision boundaries corresponding to the correct response (upper boundary) and the incorrect response (lower boundary). Applying the DDM to trial-level performance data (response times and accuracies) yields four parameters representing distinct cognitive processes: (1) **drift rate**: Average rate of evidence accumulation across trials; measuring the efficiency of information processing. (2) **boundary separation**: The distance between the two decision boundaries; measuring response caution. (3) **Non-decision time**: A combination of the preparation speed (e.g., stimulus encoding) and response execution speed (e.g., time spent pressing the button); measuring speed of non-decision processes. **starting point**: The starting point of the evidence accumulation process; measuring response bias (not considered here).  Figure adopted from Vermeent et al. (2024), licensed under CC BY 4.0.
knitr::include_graphics("figures/fig1.png")
```

# Separating EF ability from general processing speed

Recent studies show that a single general factor accounts for most of the variance in drift rates across multiple tasks [@lerche_2020; @loffler_2024; @weigard_2021].
This general factor appears to be largely stable across time [@schubert_2016; @weigard_2021].
It arises also in analyses of tasks thought to measure different abilities [@schmiedek_2007; @schmitz_2016].
Even if all tasks are designed to measure the same EF ability [@hedge_2021], variance in the general factor is fully explained by general processing speed instead of any specific EF ability [@loffler_2024].
These findings suggest that performance on standard EF tasks largely reflects basic speed of processing, rather than EF abilities.

Recent findings suggest that the effects of adversity are more general than specific.
For instance, youth with lower SES and more exposure to adversity show lower task-general raw performance scores across measures of EF, memory, and intelligence, while task-specific performance (after accounting for general variance) is often practically equivalent to zero [@bignardi_2024; @young_2024].
In addition, youth with more exposure to household threat showed lower task-general processing speed (based on drift rates), while differences in task-specific drift rates were practically equivalent to zero [@vermeent_2024a].
These findings suggest that previous research may have overestimated the associations between adversity and specific abilities.

A consequence of both sources of measurement impurity is that estimating specific EF abilities requires latent variable models including two or more measures of the same ability; yet, most studies in adversity research include only one task per EF ability.
This makes this work less suitable to investigate whether adversity is associated with specific EF abilities after accounting for general processes.
This is important for both deficit-based and adaptation-based interventions, which aim to remediate or leverage impaired or enhanced abilities, respectively. Moreover, while deficit frameworks predict impairments in both specific abilities and general processes [@vermeent_2024a], adaptation frameworks predict that specific types of adversity lead to adaptations in specific EF abilities.
For example, it has been proposed that unpredictable and dangerous environments may positively affect the ability to rapidly shift attention, promoting the detection of sudden threats and seizing fleeting opportunities [@frankenhuis_2020; @mittal_2015].
In the current research, therefore, we investigate the associations between adversity and two latent EF abilities—--inhibition and attention shifting—--after accounting for task-general processing speed.
We will use multiple tasks for each ability to obtain better estimates of specific abilities.

@@

Using a combination of DDM and structural equation modeling, we address three central questions.
First, what is the association of adversity exposure in adulthood with general processing speed that is contributing to performance across all cognitive tasks?
Second, what is the association of adversity exposure in adulthood with inhibition and attention-shifting abilities after accounting for general processing speed?
Third, what is the association of adversity exposure in adulthood with general and/or EF-specific response caution?
We focus on two types of adversity: material deprivation (a lack of access to material resources) and threat (the potential for harm imposed by others).
In previous research, both types of adversity were associated with performance on tasks used to measure EF [e.g., @fields_2021; @schafer_2022; @sheridan_2022; @vermeent_2024a; @young_2022].
In addition, both are central to recent dimensional models of adversity [@mclaughlin_2021; @sheridan_2014].

We preregistered predictions of deficit and adaptation frameworks.
Deficit frameworks predict negative associations between adversity and both the general speed of processing as well as specific abilities.
Adaptation frameworks predict intact or enhanced attention shifting ability, but not intact or enhanced inhibition ability.
We only interpret intact attention shifting ability as an adaptation if inhibition ability is also lower.
Intact performance on both abilities, however, would not provide sufficient evidence of adaptation, as it might instead suggest no association between adversity exposure and these abilities.
In line with previous work [@vermeent_2024a], we further predict that threat exposure is associated with higher task-general response caution.

@@

We tested a total of `r exclusions$sample$finish` participants from the Dutch Longitudinal Internet Studies for the Social Sciences (LISS) panel [@scherpenzeel_2011].
The LISS panel is an invitation-only, representative sample of the Dutch population consisting of approximately 7,500 individuals across 5,000 households.
LISS participants complete a yearly core battery of questionnaires about various domains of life, including one's financial situation over the past year.
In addition, LISS participants have the option to participate in further monthly studies on different topics.
We estimated power based on @kretzschmar_2019.
With $\alpha$ = 0.05 and assuming moderate reliability of our cognitive measures, we would have > 90% power to detect small effect sizes ($\beta$ = 0.1) with a sample size ranging between N = 730 and N = 980.
Therefore, we aimed for a total sample size of N = 1,000.

The current study took place between May and August 2024.
Participants were able to complete the cognitive tasks during two or more sessions to increase participation rates.
People were eligible for the study if they were between 18 and 55 years old, and had agreed to linking their LISS data to government microdata (not relevant here, but for a different study).
We excluded participants who did not have data on any cognitive task.
The final sample after exclusions consisted of `r exclusions$sample$final` participants (see Table 1).

```{r}
#| tab.id: table1
#| results: markup
table1 |> 
  flextable::autofit() |> 
  add_header_row(
    values = " ",
    colwidths = 2
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 1. "), "Descriptive statistics for the final sample."),
    part = "header"
  ) |> 
  border_remove() |> 
  border(i = 1, border.bottom = fp_border_default(), part = "header") |>  
  border(i = 2, border.top = fp_border_default(), part = "header") |>  
  border(i = 2, border.bottom = fp_border_default(), part = "header") |> 
  border(i = 14, border.bottom = fp_border_default(), part = "body") |> 
  autofit()
```

@@

We preregistered our approach for computing adversity composite scores based on observed correlations between measures.
If all measures of an adversity type (i.e., material deprivation and threat) correlated .60 or higher (indicating "strong" correlations, as was the case for material deprivation in adulthood), we calculated a uniformly weighted average.
If one or more correlations were lower than .60 (as was the case for neighborhood threat in adulthood), we applied Principal Component Analysis to the separate measures and extracted only the first principal component score.
See section 1 of the supplemental materials for frequency distributions of all adversity measures.

### Neighborhood threat in adulthood

Following our preregistration, we measured exposure to neighborhood violence in adulthood using two measures of perceived neighborhood crime, and one measure of crime victimization. 
Participants completed the Neighborhood Violence Scale [@frankenhuis_2018; @frankenhuis_2020].
This scale includes seven items about current perceived neighborhood threat (e.g., “Where I live, it is important to be able to defend yourself against physical harm”), on a scale of 1 ("Completely disagree") to 7 ("Completely agree"). 
These items were averaged and standardized.
We also included four items on perceived neighborhood threat from the LISS archive (six waves: <https://doi.org/10.17026/dans-zch-j8xt>).
Participants reported how frequently they: (1) “avoid certain areas in your place of residence because you perceive them as unsafe”, (2) “do not respond to a call at the door because you feel that it is unsafe”, (3) “leave valuable items at home to avoid theft or robbery in the street?”, (4) “make a detour, by car or on foot, to avoid unsafe areas?”.
The scale of these four items ranged from 1 (“(Almost) never”), 2 (“Sometimes”), to 3 (“Often”).
We intended to include these items again in the current study, but due to an oversight this did not happen.
This meant that we did not have data on these items for participants who never participated in any of the six waves described above (missing N = `r exclusions$sample$missing_crime_vict`). 
We decided to conduct all analyses involving threat based only on this subset of the data (N = `r exclusions$sample$final - exclusions$sample$missing_crime_vict`), because *a priori* power remained above 80% [@kretzschmar_2019].
We summed the items within each wave, and then calculated an average across waves for which participants had data.

We computed crime victimization in adulthood based on seven items from the LISS archive (six waves: <https://doi.org/10.17026/dans-zch-j8xt>) in which participants reported whether or not they had been the victim of seven types of crime in the last two years: (1) burglary or attempted burglary; (2) theft from their car; (3) theft of their wallet or purse, handbag, or other personal possession; (4) wreckage of their car or other private property; (5) intimidation by any other means; (6) maltreatment of such serious nature that it required medical attention; and (7) maltreatment that did not require medical attention. 
We also included these items again in the current study, with people answering them about the last two years.
We computed the total number of distinct crimes that participants were exposed to at any moment in time [a 'variety score'\; @sweeten_2012].

The correlations between the two measures of perceived neighborhood crime and crime victimization were low (see Table 2).
Therefore, following our preregistration, we used Principal Component Analysis to extract the first principal component score.
This score accounted for 23 % of the variance in the three measures.

### Material deprivation in adulthood

We derived measures of material deprivation in adulthood from the LISS archive, using the yearly recurring core study on household and personal income (16 waves: <https://doi.org/10.57990/1gr4-bf42>).
First, participants reported how difficult it currently is to live off the income of their household, on a scale from 0 (very hard) to 10 (very easy). 
Second, participants reported which of the following statements best described their current financial situation: (1) “we are accumulating debt”; (2) “we are somewhat eating into savings”; (3) “we are just managing to make ends meet”; (4) “we have a little bit of money to spare”; (5) “we have a lot of money to spare”. 
Third, participants reported which of the following applied to their current financial situation (0 = no, 1 = yes): (1) “having trouble making ends meet”; (2) "unable to quickly replace things that break”; (3) “having to lend money for necessary expenditures”; (4) “running behind in paying rent/mortgage or general utilities”; (5) “debt collector/bailiff at the door in the last month”; and (6) “received financial support from family or friends in the last month”.
We recoded responses so that higher scores indicated more perceived scarcity.

We first reverse-coded and averaged each measure separately across waves, and then scaled them.
As all item correlations were > .60 (see Table 2), we computed a uniformly weighted average.

::: {.landscape}

```{r}
#| tab.id: table2
#| results: markup
table2 |> 
  flextable() |> 
  flextable::set_header_labels(Variable = "") |> 
  flextable::autofit() |> 
  border_remove() %>% 
 # border(i = 1, border.top = fp_border_default(), part = "header") %>% 
  border(i = 1, border.bottom = fp_border_default(), part = "header") %>% 
  border(i = 18, border.bottom = fp_border_default(), part = "body") %>% 
  add_header_row(
    values = " ",
    colwidths = 13
  ) |> 
  flextable::compose(
    i = 1, j = 1, 
    as_paragraph(as_b("Table 2. "), "Spearman correlations between the main independent variables and covariates."),
    part = "header"
  ) |> 
  add_footer_row(
    values = " ",
    colwidths = 13
  ) |> 
  autofit()
```

:::

@@

We programmed six cognitive tasks in jsPsych 7.3 [@deleeuw_2015]: two inhibition tasks, three attention shifting tasks, and one basic processing speed task.
At the start of the session, participants entered fullscreen mode to avoid distractions from other browser tabs.
The tasks were presented against a light-gray background.
See Section 2 of the supplemental materials for information on condition manipulation checks, split-half reliability estimates, and bivariate correlations between tasks.

*Flanker Task.* This task measures inhibition of distractor interference [@eriksen_1974].
On each trial, participants saw five arrows side-by-side horizontally, pointing either left or right.
Their task was to indicate the direction of the central arrow.
The arrows were randomly presented 300 pixels above or below the center of the screen.
On 50% of the trials, all arrows pointed in the same direction (congruent trials).
On the other half, the arrows surrounding the central arrow pointed in the opposite direction (incongruent trials).
Participants first completed eight practice trials, followed by two test blocks of 32 trials each, for a total of 64 trials.
